---
title: "Infrastructure Gap"
author: "Paul Mukoki"
date: "8/6/2021"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    code_folding: hide
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1,T2A]{fontenc}
- \usepackage{amsmath}
- \usepackage{dcolumn}
mainfont: palatino
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

## Download packages if not available
if(!require(pacman)){
  
  install.packages("pacman")
  
}


## All the required packages
pacman::p_load("tidyverse", "janitor", "kableExtra", 
                      
                      "knitr", "rmarkdown", "plm", "stargazer",
               
               "skimr", "GGally", "kableExtra", "corrplot", "plm", 
               
               "punitroots")

options(scipen = 999)
```

## Preliminary: Read in the data
Here I load the dataset

```{r}
mukoki_gap_data <- readxl::read_xlsx("dataset_40_countries.xlsx") %>% 
  
  janitor::clean_names() %>% 
  
  data.frame()
```

## Dependent and Independent Variables

I select the dependent and independent variables

```{r}
modelling_data <- mukoki_gap_data %>% 
  
  select(countries, country, state, year, 
         
         income_group_2009, income_2009,
         
         overall_index, infra_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap, overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  tibble()
```

\newpage

\blandscape

### Dependent Variables

```{r}

modelling_data %>% 
  
  select(income_2009,
         
         overall_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap) %>% 
  
  na.omit() %>% 
  
  skim_without_charts() %>% 
  
  select(-n_missing, -complete_rate, -skim_type) %>% 
  
  kbl(., booktabs = TRUE, caption = "Independent Variables") %>% 
  
  kable_classic(full_width = TRUE)
  
  
```

\elandscape

\newpage


\newpage

\blandscape

```{r, fig.width = 12, fig.height = 8}
modelling_data %>% 
  
  select(income_2009,
         
         overall_index, infra_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap) %>% 
  
  na.omit() %>% 
  
  ggpairs()
```

\elandscape

\newpage


### Independent Variables

\newpage

\blandscape

We examine the independent variables. Please confirm that these are correct before I go ahead and run the regressions. I am especially keen on overall index and its meaning. What variable in the data stands for `Public-Private Partnerships (PPP) investment in Aggregate/overall infrastructure`? I dont think its the overall index. 


```{r}

modelling_data %>% 
  
  select(infra_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  skim_without_charts() %>% 
  
  select(-n_missing, -complete_rate, -skim_type) %>% 
  
  kbl(., booktabs = TRUE, caption = "Independent Variables") %>% 
  
  kable_classic(full_width = TRUE)
  
  
```

\elandscape

\newpage

```{r}
modelling_data %>% 
  
  select(infra_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  drop_na(percent_change_in_gfcf_public) %>% 
  
  cor() %>% 
  
  corrplot(method = "number", type = "lower", 
           
           diag = FALSE, number.digits = 2, 
           
           title = "Correlation Matrix for Dependent Variables")
```

\newpage

\blandscape

```{r, fig.width = 12, fig.height = 8}
modelling_data %>% 
  
  select(overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  drop_na(percent_change_in_gfcf_public) %>% 
  
  ggpairs()
```

\elandscape

\newpage


## The Fixed Effects Model
Write a regression function to run fixed and random effects  models 

```{r}
model_fixed <- plm(overall_index ~ transport_invest + electricity_invest +
         
         ict_invest + water_invest + water_invest +
         
         stock_mkt_cap + human_di + exchange_rate + 
         
         rinterestrate + percent_change_in_gfcf_public + 
         
         fdi_gdp + govenance_index, data = modelling_data, 
    
    effect = "twoways", model = "within", index = c("country", "year"))

summary(model_fixed)
```

## The Random Effects Model

```{r}
model_random <- plm(overall_index ~ transport_invest + electricity_invest +
         
         ict_invest + water_invest + water_invest +
         
         stock_mkt_cap + human_di + exchange_rate + 
         
         rinterestrate + percent_change_in_gfcf_public + 
         
         fdi_gdp + govenance_index, data = modelling_data, 
    
    effect = "twoways", model = "random", index = c("country", "year"))

summary(model_random)
```

### The Hausmann Test
Given that the test is significant, then we should use the fixed effects model .

```{r}
phtest(model_fixed, model_random)
```

## Cross-Sectional Dependence

The model shows the presence of cross sectional dependence. Thus, the unit roots tests have to consider the presence of cross sectional dependence in the panels. 

```{r}
pcdtest(model_fixed, method = "cd")
```


## Panel unit root tests in the Presence of Cross-Sectional Dependence

The main unit root tests implemented in  R's plm and punitroots, namely the tests proposed in Maddala and Wu (1999), Choi (2001), Levin, Lin, and Chu (2002), Im, Pesaran, and Shin (2003) (plm); Deme-
trescu, Hassler, and Tarcolea (2006), Hanck (2008), Costantini and Lupi (2011) (punitroots).
The first four tests assume that the series under scrutiny are cross-sectionally independent:
the last three allow for cross-dependence across the panel units.

```{r}
infra_index_wide <- modelling_data %>% 
  
  select(countries, year, infra_index) %>% 
  
  pivot_wider(names_from = "countries", values_from = "infra_index") %>% 
  
  as.matrix()
```


```{r}
my_matrix_csd <- as.matrix(modelling_data[, c("transport_invest", "electricity_invest", "ict_invest", "water_invest", "water_invest", "stock_mkt_cap", "human_di", "exchange_rate", "rinterestrate",  "percent_change_in_gfcf_public", "fdi_gdp", "govenance_index")], nrow = 640, ncol = 12)

punitroots::pCADFtest(Y = infra_index_wide[,-1], type = "drift", max.lag.y = 5, criterion = "AIC", crosscorr=0.10)
```

