---
title: "Infrastructure Gap"
author: "Paul Mukoki"
date: "8/6/2021"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    code_folding: hide
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1,T2A]{fontenc}
- \usepackage{amsmath}
- \usepackage{dcolumn}
mainfont: palatino
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

## Download packages if not available
if(!require(pacman)){
  
  install.packages("pacman")
  
}


## All the required packages
pacman::p_load("tidyverse", "janitor", "kableExtra", 
                      
                      "knitr", "rmarkdown", "plm", "stargazer",
               
               "skimr", "GGally", "kableExtra", "corrplot")

options(scipen = 999)
```

## Preliminary: Read in the data
Here I load the dataset

```{r}
mukoki_gap_data <- readxl::read_xlsx("dataset_40_countries.xlsx") %>% 
  
  janitor::clean_names() %>% 
  
  data.frame()
```

## Dependent and Independent Variables

I select the dependent and independent variables

```{r}
modelling_data <- mukoki_gap_data %>% 
  
  select(countries, country, state, year, 
         
         income_group_2009, income_2009,
         
         overall_index, infra_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap, overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  tibble()
```

\newpage

\blandscape

### Dependent Variables

```{r}

modelling_data %>% 
  
  select(income_2009,
         
         overall_index, infra_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap) %>% 
  
  na.omit() %>% 
  
  skim_without_charts() %>% 
  
  select(-n_missing, -complete_rate, -skim_type) %>% 
  
  kbl(., booktabs = TRUE, caption = "Independent Variables") %>% 
  
  kable_classic(full_width = TRUE)
  
  
```

\elandscape

\newpage


\newpage

\blandscape

```{r, fig.width = 12, fig.height = 8}
modelling_data %>% 
  
  select(income_2009,
         
         overall_index, infra_index, 
         
         transport_gap, electricity_gap,
         
         ict_gap, water_gap) %>% 
  
  na.omit() %>% 
  
  ggpairs()
```

\elandscape

\newpage


### Independent Variables

\newpage

\blandscape

We examine the independent variables. Please confirm that these are correct before I go ahead and run the regressions. I am especially keen on overall index and its meaning. What variable in the data stands for `Public-Private Partnerships (PPP) investment in Aggregate/overall infrastructure`? I dont think its the overall index. 


```{r}

modelling_data %>% 
  
  select(overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  skim_without_charts() %>% 
  
  select(-n_missing, -complete_rate, -skim_type) %>% 
  
  kbl(., booktabs = TRUE, caption = "Independent Variables") %>% 
  
  kable_classic(full_width = TRUE)
  
  
```

\elandscape

\newpage

```{r}
modelling_data %>% 
  
  select(overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  drop_na(percent_change_in_gfcf_public) %>% 
  
  cor() %>% 
  
  corrplot(method = "number", type = "lower", 
           
           diag = FALSE, number.digits = 2, 
           
           title = "Correlation Matrix for Dependent Variables")
```

\newpage

\blandscape

```{r, fig.width = 12, fig.height = 8}
modelling_data %>% 
  
  select(overall_index,
         
         transport_invest, electricity_invest,
         
         ict_invest, water_invest, water_invest,
         
         stock_mkt_cap, human_di, exchange_rate, 
         
         rinterestrate, percent_change_in_gfcf_public, 
         
         fdi_gdp, govenance_index) %>% 
  
  drop_na(percent_change_in_gfcf_public) %>% 
  
  ggpairs()
```

\elandscape

\newpage

Write a regression function to run fixed and random effects  models 

```{r}

```

