---
title: "Incomplete Results, Year 3"
author: "John Karuitha, PhD"
format: pdf
editor: visual
execute: 
  echo: false
  warning: false
  message: false
---

# **Issue**

The following students have incomplete results. If you are on the list, could you please reach me on 0736917717 and clarify?

If you did not sit for either CAT or FINAL EXAM or did not sit for both CAT and FINAL exam, just SMS me with that information (**and your admission number**), no need to call. Craft a message as follows;

         
>***Hi, I am **YOUR NAME** admission number **YOUR ADMISSION NUMBER**. I did NOT sit for CATs and exams in the following subjects.*** 


But if it is indicated that you did not sit for a CAT when you know you wrote it please call. The same case applies for the final exams.

See next page for the listing. Note that I have listed them by area of specialization.

```{r}
if(!require(pacman)){
  install.packages('pacman')
}

p_load(tidyverse, janitor, readxl, rio, kableExtra)
```

\newpage

```{r}
read_csv("bis.csv") %>% 
  kableExtra::kbl(caption = "Business Information Systems") %>% 
  kable_classic(full_width = FALSE,
                latex_options = 'hold_position')
```

```{r}
read_csv("accounting.csv") %>% 
  kableExtra::kbl(caption = "Accounting") %>% 
  kable_classic(full_width = FALSE,
                latex_options = 'hold_position')
```

\newpage

```{r}
read_csv("finance.csv") %>% 
  kableExtra::kbl(caption = "Finance") %>% 
  kable_classic(full_width = FALSE,
                latex_options = 'hold_position')
```
