---
title: "Introduction"
author: "Bahman Rostami-Tabar"
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7
    fig_height: 3.5
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE
)
library(fpp3)
options(digits = 4, width = 55)
```

## Instructor: Bahman Rostami-Tabar

- Associate Professor of Business Analytics & Management Science, Cardiff University
- Founder \& Chair, Forecasting for Social Good, International Institute of Forecasters
- Director, Data Lab for Social Good, Cardiff University

\placefig{0.6}{5.3}{width=4cm}{Bahman}

\begin{textblock}{6.5}(5.6,5.5)
\begin{alertblock}{Find me at ...}
\href{https://twitter.com/Bahman_R_T}{\faicon{twitter} @Bahman R T}

\href{https://www.linkedin.com/in/bahman-rostami-tabar-1046171a/}{\faicon{linkedin}  Bahman Rostami-Tabar}

\href{https://bahmanrt.com}{\faicon{home} bahmanrt.com}

\href{mailto:rostami-tabarb@cardiff.ac.uk}{\faicon{envelope}  rostami-tabarb@cardiff.ac.uk}
\end{alertblock}
\end{textblock}
\vspace*{10cm}

## International Institute of Forecasters

- a nonprofit organization founded in 1982, is dedicated to developing and furthering the generation, distribution, and use of knowledge on forecasting

- [Forecasting for Social Good](https://forecasters.org/programs/communities/forecasting-for-social-good-f4sg/)

## Assumptions

  * This is not an introduction to R. We assume you are broadly comfortable with R code, the RStudio environment and the tidyverse.
  * This is not a statistics course. We assume you are familiar with concepts such as the mean, standard deviation, quantiles, regression, normal distribution, likelihood, etc.
  * This is not a theory course. We are not going to derive anything. We will teach you time series and forecasting tools, when to use them, and how to use them most effectively.

## Approximate outline of the workshop

\centering\fontsize{11}{11}\sf
\begin{tabular}{rp{6.6cm}l}
  \bf Session & \bf Topic  \\ 
  \midrule
  1       & Introduction \\
  2       & Prepare data (tidy) \\
  3       & Time series patterns and graphics \\
  4       & Time series features \\
  5       & Introduction to forecasting \\
  6       & Forecaster’s toolbox-part 1 \\
  7       & Forecaster’s toolbox- part 2 \\
  8       & Exponential smoothing models \\
  9       & ARIMA models    \\
  10       & Time series regression  \\
  11      & Dynamic regression   \\

\end{tabular}

##  Tidy workflow

```{r workflow, echo = FALSE}
line_curve <- function(x, y, xend, yend, ...){
  geom_curve(
    aes(x = x, y = y, xend = xend, yend = yend),
    arrow = arrow(type = "closed", length = unit(0.03, "npc")),
    ...
  )
}

ggplot() +
  geom_text(
    aes(x = x, y = y, label = label),
    data = tribble(
      ~ x, ~ y, ~ label,
      1, 0, "Tidy",
      7/3, 0, "Visualise",
      3, 0.5, "Specify",
      11/3, 0, "Estimate",
      3, -0.5, "Evaluate",
      5, 0, "Forecast"
    ),
    size = 5
  ) +
  geom_segment(
    aes(x = x, y = y, xend = xend, yend = yend),
    data = tribble(
      ~ x, ~ y, ~ xend, ~ yend,
      1.3, 0, 1.9, 0,
      4.1, 0, 4.6, 0
    ),
    arrow = arrow(type = "closed", length = unit(0.03, "npc"))
  ) +
  line_curve(7/3, 0.1, 8/3, 0.5, angle = 250, curvature = -0.3) +
  line_curve(10/3, 0.5, 11/3, 0.1, angle = 250, curvature = -0.3) +
  line_curve(8/3, -0.5, 7/3, -0.1, angle = 250, curvature = -0.3) +
  line_curve(11/3, -0.1, 10/3, -0.5, angle = 250, curvature = -0.3) +
  theme_void() +
  xlim(0.8, 5.2) +
  ylim(-0.6, 0.6) +
  coord_equal(ratio = 1)
```

## Key reference

\large

\begin{block}{}\bf
\hangafter=1\hangindent=.3cm
 {Hyndman, R.~J. \& Athanasopoulos, G. (2022) \emph{Forecasting: principles and practice}, 3rd ed.}
\end{block}\pause
\begin{alertblock}{}\Large
\centerline{\bf OTexts.org/fpp3/}
\end{alertblock}

\pause

- Recommended

    - [Profit From Your Forecasting Software](https://amzn.to/36UKZ96)
    - [Principles of Business Forecasting](https://amzn.to/2UyBMkz)
    - [Managerial Demand forecasting](will be published in March 2023)


## Install required packages

Please bring your own laptop and install the latest version of [R](https://cran.ma.imperial.ac.uk/) and [Rstudio](https://rstudio.com/products/rstudio/download/#download)


Install packages

```{r,eval=FALSE, echo=TRUE}
library(tidyverse)
library(fpp3)
library(GGally)
library(sugrrants)
library(astsa)
library(dygraphs)
library(xts)
library(ggthemes)
```

## Schedule for each day

| Time          | Activity         |
| :------------ | :--------------- |
| 09:00 - 10:30 | Session 1        |
| 10:30 - 11:00 | *Coffee break*   |
| 11:00 - 12:30 | Session 2        |
| 12:30 - 13:30 | *Lunch break*    |
| 13:30 - 15:00 | Session 3        |
| 15:00 - 15:30 | *Coffee break*   |
| 15:30 - 17:00 | Session 4        |


## Poll: How experienced are you in forecasting and R

Please complete the pools:

[Pool](https://www.menti.com/al8a69cp3dkh)

