---
title: "**Tidy Time Series Analysis**"
subtitle: "*Final Day Project*"
author: 
    - name: John Karuitha
      url: www.linkedin.com/in/Karuitha
      affiliation: "Karatina University, Kenya"
      affiliation-url: www.karu.ac.ke
description: "Independent Data Analysis Project"
date: today
title-block-banner: true
title-block-banner-color: "black"
echo: true
message: false
warning: false
format: 
    html:
        theme: 
            light: flatly
            dark: darkly
        number-sections: true
        code-fold: true
        code-background: true
        toc: true
        toc-title: "Contents"
        toc-depth: 3
        toc-float: true
        linkcolor: "blue"
        tocolor: "blue"
        link-citations: true
    pdf:
        header-includes: |
            \usepackage{pdflscape}
            \usepackage[OT1]{fontenc}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
        toc: true
        toc-depth: 3
        toc-title: "Table of Contents"
        linkcolor: blue
        toccolor: blue
        number-sections: true
        number-depth: 3
        documentclass: report #can always change doc type to your choice
        margin-left: 30mm
        margin-right: 30mm
        link-citations: true
editor: visual
bibliography: ref.bib
---

# **Background**

```{r}
if(!require(pacman)){
        install.packages("pacman")
        library(pacman)
}

pacman::p_load(tidyverse, fpp3, lubridate, 
tsbox, dygraphs)
```

@hyndman2018forecasting

# **Data**
In this section, I read in the data, summarise and visualize the data. 

## Data and Data Sources

The data is an estimate of monthly Malaria cases as collected by the Ministry of Health in Kenya.

```{r}
## Read in the data ----


```

The data consists of the following variables. 

```{r}
## Desciption of the variables 
tribble(~ Variable, ~ Description)

arty_farty
```

## Data Visualization



# **Model Development**




# **Model Evaluation**



# **Concluding Remarks**



# **References** {.unnumbered}


