---
title: "Analysis of Men Olympics Marathon Data from IOC"
subtitle: "Which Countries Do Better in Marathons?"
author: "John Karuitha"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)

if(!require(pacman)){
        
        install.packages("pacman")
}

pacman::p_load(tidyverse, rvest, countrycode, 
               
               data.table, glue, lubridate,
               
               readxl, tidyquant, kableExtra, knitr)
```

## Background

The `Marathon`, one of the initial olympics events covers a distance of 42.195 km. As one of the highlights of modern olympics, the race attracts millions of fans globally. In this article, I examine the performance of male athletes in olympics marathons. Specifically, I examine;

- Which countries have produced the highest number of olympics male marathon winners.
- What is the prime age for a marathoner.

## Data

I use data from the International Olympics Committee (IOC) from 1950 to 2022. I scrapped the data from the IOC [website](https://www.worldathletics.org/records/all-time-toplists/road-running/marathon/) on this link <https://www.worldathletics.org/records/all-time-toplists/road-running/marathon/>. 

```{r, echo = FALSE}
read_chunk("marathon_script.R")
```

```{r theme_set}

```

The data spans 48 web pages. I used the function below to scrap the data. 

```{r scrapper}

```

The scrapping takes considerable time. Hence, I have saved the data in an excel sheet to save time. However, you can always scrap the data. 

Here, I read in the data from the excel file. 

```{r read_data}

```

The data has `r nrow(final_marathon_data)` rows and `r ncol(final_marathon_data)` columns.

```{r head_str}

```


```{r map_visual}

```

