---
title: "Most Popular R packages on Tweeter"
author: "John Karuitha"
date: "10/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(rtweet)
library(tidyverse)
library(rvest)
library(tidytext)
```
## Background
In this mini-project, I extract data from tweeter regarding popular `R` packages. The aim is to get a picture of the projects 
## The List of R packages
I get a list of R packages from <https://cran.r-project.org/>. 

```{r}
my_r_packages <- read_html("https://cran.r-project.org/web/packages/available_packages_by_name.html") %>% 
        ## Capture the nodes for the names
        html_nodes("td:nth-child(1)") %>% 
        
        html_text()

head(my_r_packages)

available.packages() %>% rownames() %>% length()
```

## Tweets Mentioning R and Rstats

Next, I get tweets that mention `#rstats`. 

```{r}
df_tweets <- search_tweets(q = "#rstats", n = 10000, include_rts = FALSE)

head(df_tweets$description)
```

I then `unnest` all the tokens to create a new column `words` from the `text` column of tweets. After removing stopwords. Stopwords are essential in communication but do not carry meaning on their own, for instance prepositions. Next, I filter for names of `R` packages and then count the words, sorted in descending order of frequency. 

```{r}
df_tweets %>% 
        
        unnest_tokens(word, text) %>% 
        
        anti_join(stop_words) %>% 
        
        filter(word %in% my_r_packages) %>% 
        
        count(word) %>% 
        
        arrange(desc(word))
```



