---
title: "Time series analysis and forecasting"
output: 
  word_document: 
    toc: yes
  html_document: 
    toc: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
a
```

# Prepare data

## Read data

```{r read-data}

```

## Create a tsibble

```{r tsibble}
#check and fix duplicates


# Create tsibble using as_tsibble

#Check gaps in time and fix it

```

# Visualise

## time plot

```{r time-plot}

```

## seasonal and subseries


```{r seasonal-plot}

```

## autocorrelation

```{r autocorrelation-plot}

```

# Specify and train models

## Specify and train models using time series cross validation


## Split

```{r split-data}

```


## Specify and train models using time series cross validation


```{r specify-train}

```


## Generate forecasting



```{r forecast}


```


# Evaluate forecast accuracy

## Evaluate point forecast accuracy

```{r point-accuracy, options}

```

## Evaluate prediction interval forecast accuracy


```{r interval-accuracy, options}

```


# Check residuals for the selected model


```{r residual, options}

```


# Forecast for the future using the selected model and visualise forecast


```{r forecast-visualise, options}

```

