---
title: "Purchase of Office Equipment"
subtitle: "From Which Outlet Should We Buy Office Equipment?"
author: "John Karuitha"
date: "`r format(Sys.Date(), format = '%A %B %d, %Y')`"
output: 
  html_document:
    fig_height: 3
    fig_width: 5
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE, message = FALSE, 
                      
                      warning = FALSE)
## Install and load package manager
## NB: This section requires an internet connection

if(!require(pacman)){
        
        install.packages("pacman")
        
}

## Load required packages, installing them if not available
pacman::p_load(tidyverse, ggthemes, GGally, janitor, skimr, kableExtra, 
               
               huxtable, flextable, knitr, rmarkdown, officedown, 
               
               officer, corrplot, dendextend, flexclust, NbClust,
               
               here, questionr, ggmosaic, OddsPlotty, tidymodels,
               
               lift, data.table,  dfidx, mlogit, lmtest, officedown,
               
               markdown, devtools, remotes, mosaic, thematic
               
               )

## Install bbplot
devtools::install_github('bbc/bbplot')
## Set theme for plots
theme_set(bbplot::bbc_style())

## Set number of significant digits
options(digits = 2)

knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Background

When procuring goods and services, there are many options from the variety of the goods and services to the different vendors. In this project, we examine the choice of vendor. We use data from a survey of 200 respondents regarding the importance of several store attributes when choosing where to buy office equipment.

## Objectives 


## Data

```{r}
## Read the data
office <- read_csv("office (2).csv") %>% 
        
        clean_names()

```



```{r, eval = FALSE}
## Examine first six rows of the data.
office %>% 
        
        head() %>% 
        
        kbl(., caption = "First Six Rows of the Data") %>% 
        
        kable_classic(full_width = TRUE)
```

## Session Information

I owe this report to the following R packages. My appreciation to the R Foundation, R Studio and the Free Software Movement. 
  
```{r echo=FALSE}
sessionInfo()  # could use devtools::session_info() if you prefer that
```
  