---
title: "exponential_smoothing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# DO NOT FORGET TO LOAD PACKAGES
library(tidyverse)
library(fpp3)
admission_daily <- read_csv(
  file = "data/ae_uk.csv") |> mutate(arrival_time= dmy_hm(arrival_time)) |> 
  dplyr::distinct() |> as_tsibble(index = arrival_time, key = c(gender,type_injury), regular=FALSE) |> index_by(date = lubridate::as_date(arrival_time)) |> 
  summarise(arrival=n(), .groups = "drop") |> fill_gaps(arrival=0)
```

# Lab session 7: EXPONENTIAL SMOOTHING

Fitting ETS model (E: Error, T:Trend, S: Seasonality)
Remember the following from ETS fucntion

N: None (No trend, no seasonality)
A  additive
Ad: additive damped
M: multiplicative

The function for ets models in fable is `ETS()`
check ?ETS in your Console

If we don't provide arguments inside ETS(), then it is an automatic ETS! It will examine different models and return the one with lowest AICc.


```{r fit-ets}
admission_ets <- admission_daily %>%
  model(
    automatic_ets = ETS(arrival)
  )
```

let's now observe the fitted model

```{r label, options}
admission_ets %>% report()# provide a nice output of models with parameters estimated
admission_ets %>% tidy()# provide parameters and its estimation

admission_daily %>% head() %>% mutate(wday=wday(date, label = T))
```

We can use the `components()` function to observe the values corresponding to level, trend, seasonal equations in ETS framework. Each column corresponds to one equation:

```{r observe-component}
admission_ets %>% components()# provide a nice output of models with parameters estimated
```

Now, we can forecast with the fitted ETS model:

```{r forecast-ets, options}
admission_ets %>%
  forecast(h = "42 days") 
```


# Explore models and parameters effect 

You can use the following specific functions: `error()`, `trend()`, `season()` to manulaly specify the type of pattern ("N", "A", "Ad", "M") and also their corresponding parameters

error(method = c("A", "M"))

trend(method = c("N", "A", "Ad"),
      alpha = NULL, alpha_range = c(1e-04, 0.9999),
      beta = NULL, beta_range = c(1e-04, 0.9999),
      phi = NULL, phi_range = c(0.8, 0.98))

season(method = c("N", "A", "M"), period = NULL,
       gamma = NULL, gamma_range = c(1e-04, 0.9999))

In the following R-chunk, you can change parameters and the type of pattern to see their impact on fitting and forecast
  
```{r alpha}
admission_ets <- admission_daily %>%
  model(
    `alpha = 0.05` = ETS(arrival ~ error("A") + trend("A", alpha = .05) + season("N")),
    `alpha = 0.15` = ETS(arrival ~ error("A") + trend("A", alpha = .15) + season("N")),
    `alpha = 0.5` = ETS(arrival ~ error("A") + trend("A", alpha = .5) + season("N")),
    `alpha = 0.9` = ETS(arrival ~ error("A") + trend("A", alpha = .9) + season("N"))
  )

admission_ets %>% augment() %>% filter_index("2016" ~ .) %>% 
  ggplot(aes(x=date))+
  geom_line(aes(y=arrival, colour= "Actual"))+
  geom_line(aes(y=.fitted, colour= factor(.model)))+
  ggthemes::scale_color_colorblind()+
  labs(colour ="")
  
```

You can also explore `beta` ,`phi` and `gamma`

```{r parameter-ets}
admission_ets <- admission_daily %>%
  model(
    simple_ES = ETS(value ~ error("A") + trend("N", alpha = 0.1) + season("N")),
    Holt = ETS(value ~ error("A") + trend("A", alpha = 0.1, beta = .15) + season("N")),
    holt_damped = ETS(value ~ error("A") + trend("Ad", alpha = 0.1, beta = .1, phi = .8) + season("N")),
    holt_winter = ETS(Trips ~ error("A") + trend("A", alpha = 0.1, beta = .1) + season("A", gamma = .05))
  )
```

You can visualise forecasts:

```{r forecast-viz}
fc <- admission_ets %>% forecast(h = "42 days")
fc %>%
  autoplot(filter_index(admission_daily, "2016" ~ .), level = NULL) +
  labs(title="Australian domestic tourism",
       y="Overnight trips (millions)") +
  guides(colour = guide_legend(title = "Forecast"))
```