---
title: "Analyzing the 100 Metres Men's Sprint"
runningheader: "100 Metres Men Best Times" # only for pdf output
subtitle: "100 Metres Sprint Best Times, 1900 - Present" # only for html output
author: "John King'athia Karuitha"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(tidyverse)
library(rvest)
library(janitor)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Background

I use data from [World Athletics](https://www.worldathletics.org/records/all-time-toplists/sprints/100-metres/outdoor/men/senior?regionType=world&timing=electronic&windReading=regular&page=21&bestResultsOnly=false&firstDay=1900-01-01&lastDay=2021-09-20) on the best times posted by male athletes in the 100 metres sprint from 1900  to present  ^[The data is available on this link <https://www.worldathletics.org/records/all-time-toplists/sprints/100-metres/outdoor/men/senior?regionType=world&timing=electronic&windReading=regular&page=21&bestResultsOnly=false&firstDay=1900-01-01&lastDay=2021-09-20>]. Note that many athletes appear multiple times provided the times. For instance, Usain Bolt posted top 3 best times in different races. 

```{r}

```

The data consists of the following variables;

- rank: Starting from the athlete that has posted the best time to date.
- mark: The time (in seconds) posted by the athlete.
- wind: The wind assist. Negative speeds indicate the athlete was running against the wind. 
- competitor: The name of the athlete.
- dob: Date of birth of the athlete.
- nat: Nationality of the athlete.
- pos: position of the athlete in the given race. 
- venue: Venue of the race.
- date: Date the race happened. 
- results_score: The score of the athlete in the race by World Athletics. 

