---
title: "**Analyzing `lendingclub.com` Loan Data**"
author: "John Karuitha"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
always_allow_html: true
number_sections: true
output:
  pdf_document: 
    toc: true
    toc_depth: 4
  html_document: 
    theme: yeti
    code_folding: hide
    toc: true
    toc_depth: 4
subtitle: "*What Kind of Customers Take Loans? For What Purpose?*"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1,T2A]{fontenc}
- \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(skimr)
library(tidyverse)
library(knitr)
library(GGally)
library(ggthemes)
library(tibble)
library(magrittr)
library(gghalves)
library(patchwork)

options(digits = 4)
options(scipen = 999)
theme_set(theme_clean())
```


```{r, echo = FALSE}
knitr::read_chunk("code/my_loan_script.R")
```


```{r loaddata, echo = FALSE}
```


```{r, child = "text/Background.Rmd"}

```

```{r, child = "text/Objectives.Rmd"}

```

```{r, child = "text/Data.Rmd"}

```

# Exploratory Data Analysis
In this section, I load the data and do exploratory data analysis, starting first with data visualisation followed by summary statistics aimed at uncovering the insights stated in the objectives section above. 

```{r loandatastructure}
```

## Data Visualization

### The Correlation Matrix
Figure () below shows the correlation matrix of the numeric explanatory variables and the response variable. I will refer to this correlation matrix in the analysis that follows. 

### What type of customers does the lender perceived to be risky? Are they as risky?
In this section, I use data vislualisation techniques assess the types of customers that the lender percieves to be riskiest and whether this perception is supported by the data. The interest rate is proxies the risk assessment of the customer by the lender. 

The correlation matrix in table () shows a high correlation between fico score and interest rates charged on loans (-0.71). Given that the interest rate has a very high correlation with fico score, it appears like the lender, to a large extent relies on fico scores to assess credit risk. Also, fico score has a high negative correspondence with revol_util(the amount of credit utilised relative to the total credit available) at 0.541. This observation means that individuals with high fico scores tend to utilize their lines of credit less than individuals with a low fico score. However, there are a few outliers as figure () shows. These are individuals with high fico scores and also a high loan utilisation rate. These customers could be targeted with relationship banking to offer personalised services. 



However, there is a high positive correlation between revol_util and interest rates. This observation implies that the lender views people with high loan utilisation as presenting a greater risk of default. However, as we have seen, people with high fico scores present low default risk and yet have high utilisation rate. 

Other variables that have a high 

>Take away 1: People with high fico score present a low default risk and a high rate of loan utilisation. This segment of customers has a high return to risk ratio. 

 Figure () supports this observation, showing that high fico scores correspond to lower loan delinquency. 

>Take away 2: Segment customers with high loan utilisation rate using the fico scores. Charge a lower rate to those with high scores and retain high rates to customers with lower scores. In my assessment, the lender should use a fico cut off rate of between 700 and 750 and a loan utilisation cap of 100%.  

\blandscape
```{r ggallyvisual, fig.width=9, fig.height=6, fig.cap = "Correlation Matrix for Numeric Predictors"}
```

\newpage

```{r whoisrisky, fig.width=9, fig.height=6}
```

\newpage

```{r bank_risk_perception, fig.width=9, fig.height=6}
```

\elandscape

## Summarising the Data

### Basic Modelling

```{r, child = "text/Conclusion.Rmd"}

```


